---
output: 
  github_document:
    number_sections: true
documentclass: scrartcl
linestretch: 1.5
fontfamily: palatino
fontsize: 12pt
papersize: a4
geometry: margin=1in
header-includes:
  - \usepackage{parskip} 
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage[utf8]{inputenc}
  - \usepackage{makecell}
  - \usepackage{xcolor}
bibliography: asthma.bib
csl: harvard.csl
---

``` {r include = FALSE, echo = FALSE}
## Packages
library(ggplot2)   # Plots
library(corrplot)  # Plots
library(dplyr)     # Data manipulation
library(tidyverse) # map2_df
library(broom)     # tidy
library(survey)    # Complex survey design
library(tableone)  # Survey table
library(DescTools) # Cramer's V
library(caret)     # Model fitting
library(glmnet)    # Logit
library(ROSE)      # Resampling
library(yardstick) # Model evaluation
library(knitr)     # Styling
library(kableExtra) # Table styling

## Pre-load data
df.p <- read.csv("data/primary.csv", stringsAsFactors = TRUE)
df.p <- df.p %>% select(-c(id))
df.v <- read.csv("data/weighted.csv", stringsAsFactors = TRUE)

## Random seed
set.seed(42)
```

<!-- Title page -->
\pagenumbering{roman}
\newpage
\begin{center}

\LARGE{Gender-Specific Demographic and Socieconomic Determinants of Asthma Prevalence}

\bigskip
\bigskip

\large{Niall Ryan}
\end{center}
\newpage

<!-- Abstract -->
\begin{center}
\large{Abstract}
\end{center}
\bigskip

Asthma is a chronic inflammatory disease of the lung airways. 
The morbidity and severity of asthma varies substantially across genders, ethnicities, socioeconomic backgrounds,
and geographical locations. 
This study investigates the differences in the effects of socio-demographic factors on asthma prevalence in male
and female gender groups. Using unweighted 5-cycle (10-year) samples of NHANES between 2009 and 2018,
logistic regression and random forest models are fitted for the overall population and subpopulations by gender
to identify the most significant factors and quantify the effects.
The primary analysis results suggest that gender is one of the most significant predictors of asthma prevalence 
(49.11% increase in risk for women, $CI_{95}=[44.07\%, 53.70\%]$).
Between gender groups, obesity affects male asthma prevalence stronger than female asthma prevalence 
(4.8 times increase versus 2.2 times increase). *Non-Hispanic Black* men are 80% more likely to suffer from asthma
compared to *Non-Hispanic Black* women ($CI_{95}=[52.69\%,119.35\%]$).
The risk of asthma in women with no high school education is 20% higher compared to men 
($CI_{95}=[19.37\%,21.05\%]$).
The effects of smoking on asthma prevalence are consistent between genders with only weak evidence for the
significance of smoking in either group. Poverty effects are stronger among male respondents. The results
emphasise the importance of obesity, poverty, and ethnicity as factors contributing to asthma prevalence.

\newpage 

<!-- TOC -->
\tableofcontents
\newpage

# List of tables{-}
\renewcommand{\listtablename}{} <!-- removes default section name -->
\listoftables

\newpage
# List of figures{-}
\renewcommand{\listfigurename}{}
\listoffigures

\newpage

<!-- Body -->
\pagenumbering{arabic} 

# Introduction{#intro}

Asthma is a serious health care concern worldwide, affecting millions of people and incurring substantial
social costs. The severity of the asthma prevalence issue is due to a number of factors, including the difficulty 
of diagnosis, incomplete understanding of the disease's mechanics and causes, and the varying short-term and
long-term effects across populations by geographic location and socio-demographic conditions. This study contributes
to a recent strand of literature that focuses on socio-demographic determinants of asthma, exploring a 10-year
sample of the US population. It emphasises the interaction of gender with other socio-demographic variables
and explores both the inferential and predictive power of statistical models.

The rest of this report is organised as follows. Section \ref{back} provides further information on 
the study's motivation and the underlying theoretical framework. 
Section \ref{methods} describes the data collection process and justifies
the methodological approach to modelling asthma prevalence in different populations. 
Section \ref{analysis} describes in detail the main estimations and results in the 
context of the theoretical framework and published empirical literature.
Section \ref{conc} concludes with a discussion of the study's limitations and directions for future research.

# Background{#back}

Asthma is a chronic inflammatory lung disease caused by a complex combination of genetic and environmental
factors [@chippsDeterminantsAsthmaIts2004]. It is one of the most common non-communicable diseases and a major cause
of morbidity, disability, and quality of life deterioration, affecting more than 300 million people worldwide 
[@GlobalAsthmaReport2018]. Both the mechanisms underlying asthma and its causes are not completely understood, with
recent research suggesting that the disease prevalence cannot be attributed to a single factor or a set of
non-interacting factors [@dharmageEpidemiologyAsthmaChildren2019]. Although, in general, the number of cases of 
asthma is expected to grow globally, there are significant differences in asthma prevalence between countries 
and geographic regions [@toGlobalAsthmaPrevalence2012a].

Non-equitable access to health care due to race, ethnicity, and socioeconomic disparities is often linked to
higher levels of disease prevalence, morbidity, and severity in disadvantaged communities. In the US, the number
of cases of the most common severe diseases, such as cancer or diabetes, is disproportionately larger among
the non-white population [@williamsAchievingEquityEvolving2016]. Similar findings are reported for respiratory 
diseases [@celedonRespiratoryHealthEquality2014], emphasising higher smoking incidence and rates of pollution as
major contributors to respiratory health inequities. 

At the same time, one of the suggested factors of 
asthma prevalence is urbanisation, which implies that wealth may be positively associated with the number of asthma
cases. This argument finds support in empirical studies; in general, wealthier communities are expected to have
a higher prevalence of asthma while economically disadvantaged regions tend to see more severe cases 
[@cruzSocialDeterminantsAsthma2010; @toGlobalAsthmaPrevalence2012a; @dharmageEpidemiologyAsthmaChildren2019].
Controlling for urbanisation, the evidence for higher risk of asthma dependent on socio-demographics remains 
consistent [@williamsSocialDeterminantsTaking2009]. Apart from health care access, smoking, and pollution, other
explanations for disparities based on socio-economic factors include health literacy and education, stress and 
mental health, and population clustering.

All asthma determinants discussed so far are modifiable and, thus, empirical results obtained from investigating them
can be used to directly inform health care and socio-economic equity policies. However, exploring the effects of 
non-modifiable risk factors is important as well since, e.g., genetic factors may interact with non-modifiable 
factors and alter the associated asthma outcomes 
[@chippsDeterminantsAsthmaIts2004; @martinezGenesEnvironmentsDevelopment2006]. 
Another non-modifiable factor that significantly affects asthma prevalence and severity is gender -- 
women, especially younger women, are more likely to develop asthma symptoms, and the symptoms tend to be more 
pronounced [@pignataroAsthmaGenderFemale2017]. Investigating the direct effects of non-modifiable risk 
factors and their interactions with modifiable factors provides relevant evidence for understanding the population 
differences in asthma prevalence and designing personalised intervention policies 
[@williamsAchievingEquityEvolving2016; @greenblattGenderspecificDeterminantsAsthma2017].

The National Health and Nutrition Examination Survey (NHANES) is a repeated survey with a complex design
conducted continuously, every two years, in the US. NHANES data contains a wide variety of measurements,
including health status, demographics, living conditions, dietary conditions, etc. The data are widely employed
in health care studies and educational contexts due to ease of access and consistency across time.
This study is inspired by the investigation in @greenblattGenderspecificDeterminantsAsthma2017, which observes
a sample of 3 cycles of NHANES and explores socio-demographic determinants of asthma prevalence by gender.
@greenblattGenderspecificDeterminantsAsthma2017 report large and statistically significant differences in the
adverse effects of *obesity* and *smoking* between genders. This study seeks to validate and extend 
the investigation in @greenblattGenderspecificDeterminantsAsthma2017 by employing a larger sample of NHANES data
and exploring improvements to the original methodology.

# Data and Methods{#methods}

## Data{#data}

This study combines 5 cycles (10 years) of NHANES data from 2009 to 2018 into a single sample. The dependent variable
is the current diagnosis of asthma, coded from two responses from the NHANES questionnaire: *"Has a doctor or other 
health professional ever told you that you have asthma?"* and *"Do you still have asthma?*". The respondents who 
answered "yes" to both questions are classified as having asthma; the respondents who answered "no" to the first 
question are classified as not having asthma. 
The sample also includes respondents' gender, age, ethnicity, education, BMI, 
smoking status, and poverty indicator. 

Following @greenblattGenderspecificDeterminantsAsthma2017, the poverty 
indicator is recoded into a binary variable that takes value 1 if the respondent household income is below the 
poverty line and value 0 otherwise. Likewise, the education variable is recoded with three levels -- 
below high school education, high school education, and above high school education. The *ethnicity* variable is 
transformed to combine *Mexican* and *Other Hispanic* respondents. Respondents of age below 22 are excluded from the 
sample to avoid bias in education responses. *I don't know* and *Refused to answer* responses to any of the survey 
questions are considered missing observations. 
Across all variables, for respondents of age 22 and above, the number of these responses is 142 ($N = 27,889$).
Apart from *I don't know* and *Refused to answer* responses, there is a large number of *Not available* responses
in the sample: 15,650 (56.12%), 2,906 (10.42%), and 1,467 (5.26%) for *smoke*, *poverty*, and *bmi* variables, 
respectively. Overall, 17,336 (62.16%) of respondents are excluded from the final sample; the sample size
is $N = 10,553$. Since the number of excluded observations is rather large, section \ref{expl} contains
a discussion concerned with the validity of the primary sample.
Table \ref{tbl:vars} in Appendix \ref{a1} reports NHANES variables employed in the study with descriptions.

## Primary Analysis{#prime}

For the primary analysis, the combined sample is considered a sample of independent observations that is
*not representative* of the US population. 
Since the observations are independent, we can treat the sample as a cross-sectional survey
at a single time point and conduct the analyses accordingly [@firebaughAnalyzingRepeatedSurveys1997]. 
A problem with this approach, of course, is that it assumes that the relationship between predictors 
and the asthma outcome is constant across time. 
To mitigate that, the sample includes a year dummy and the stability of observed indicators is investigated
before and during modelling.

The binary dependent variable implies a classification problem. The response classes are highly unbalanced with only
about 10.25% of the respondents having asthma. Imbalanced classification is a difficult and unsolved problem
in various applications of statistics and especially in health care studies [@heLearningImbalancedData2009].
The main problem the imbalance response presents is that any naive statistical model will be biased towards
the majority class. @brancoSurveyPredictiveModeling2016 provide an extensive review of the approaches
to mitigate the bias resulting from imbalance. 
This study combines the measurement approach and the resampling approach. With the former, models are optimised
for a metric other than accuracy -- since accuracy can be very high and, at the same time, not meaningful
if the model predicts that all observations belong to the majority class. This study uses ROC as the optimisation
metric. 
Another metric commonly employed in imbalanced 
classification is Cohen's Kappa [@mchughInterraterReliabilityKappa2012]. The choice of ROC is justified by easier
interpretation and estimation of the metric. An additional justification is, arguably, better control of the
trade-off between specificity and sensitivity. As much as it is important to correctly classify individuals
suffering from asthma, it is also important to maintain a reasonable false positive rate since false positives
may result in economic losses due to misuse of health care resources.

The study employs logistic regression and random forest for classification, 
with standardised continuous predictors and one-hot encoding for categorical predictors.
A baseline logistic model is estimated with a raw sample and compared with upsampled regression and random forest
models. The study employs Random Over-Sampling Examples (ROSE) [@lunardonROSEPackageBinary2014], which
produces additional observations in the minority class by synthetic generation based on the 
nearest neighbours approach.
The ROSE procedure is available as the *ROSE* R package. Cross-validation for resampling and model fitting
is performed using the *caret* package.
While it is possible to employ other imbalanced classification techniques, this study focuses on the most well-known.
The body of literature exploring imbalanced classification is massive [@brancoSurveyPredictiveModeling2016] and
trying out all techniques deserves a separate study.

\newpage

# Analysis{#analysis}

## Exploratory Analysis{#expl}

Table \ref{tbl:byg} shows the frequencies for all variables in the primary sample by gender. For comparison,
Table \ref{tbl:bygv} in Appendix \ref{a1} presents the same (unweighted) statistics for the full sample. As
we can see, the differences between samples are substantial -- the primary analysis sample, for instance,
is skewed towards male respondents. The fact that these differences emerge even before applying survey weights
emphasises the reason to treat the primary analysis sample as non-representative of the population.

Female respondents are more likely to be suffering from asthma in both samples, although, in the primary analysis 
sample, the difference is much more pronounced (7.7 percentage difference in the primary sample vs 1.7 percentage
difference in the full sample). 
The primary analysis sample is, predictably, biased towards higher rates of smoking due to the exclusion of
*Not available* responses to the smoking question. The average BMI is higher in both male and female respondents
in the primary analysis sample. In both samples, however, female respondents tend to have a higher BMI with larger
deviations from the mean.
Tables \ref{tbl:byy} and \ref{tbl:byyv} in Appendix \ref{a1} report asthma
prevalence counts by year and gender. In the full sample, gender differences are approximately consistent
across time. In the primary sample, we observe substantial fluctuations in the asthma prevalence ratios by gender. 
The time-dependent differences between samples are, most likely, 
caused by the exclusion of observations with missing responses. 

Table \ref{tbl:bya} reports asthma prevalence counts across all variables. The same information for the full
sample is reported in Table \ref{tbl:byav} in Appendix \ref{a1}. As is apparent from the Tables, this study ignores
the absolute majority of asthma cases among individuals younger than 22 years. 
Respondents in the *Non-Hispanic Black* demographic group have the largest ratio of asthma cases across 
ethnicities (13.07%), followed by *Non-Hispanic White* (12.45%), *Other* (10.07%), and, lastly, *Hispanic* (8.12%).
*ethnicity* ratios in the primary analysis sample are consistent with the ratios in the
full sample with an overall skew towards higher asthma prevalence in all groups.

``` {r echo = FALSE, message = FALSE, warning = FALSE}
byg <- CreateTableOne(data = df.p, includeNA = TRUE, 
                      strata = "gender", test = FALSE,
                      var = c("age", "ethnicity", "poverty",
                              "education", "bmi", "smoke",
                              "asthma"))

kableone(byg,
         format = "latex",
         escape = TRUE,
         booktabs = TRUE,
         caption="Variable frequencies by gender\\label{tbl:byg}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

Asthma prevalence in respondents with household income below the poverty line is almost twice as high compared
to wealthier households (16.15% versus 9.93%). Almost half of the respondents suffering from asthma are smokers; 
however, the differences in smoking prevalence between *Yes* and *Never* groups are not as striking as the
differences in income (12.66% versus 10.43%). Similarly, asthma prevalence in respondents with below 
high school (12.70%) and high school (11.33%) levels of education is slightly higher than among respondents with
*Above high school* education (10.78%). Again, all of these
observations generally hold in the full sample, which provides justification for using the unweighted subset
of the population to explore the determinants of asthma prevalence.

``` {r echo = FALSE, message = FALSE, warning = FALSE}
bya <- CreateTableOne(data = df.p, includeNA = TRUE, 
                      strata = "asthma", test = FALSE,
                      var = c("gender", "age", "ethnicity", "poverty",
                              "education", "bmi", "smoke"))

kableone(bya,
         format = "latex",
         escape = TRUE,
         booktabs = TRUE,
         caption="Demographic variables by asthma\\label{tbl:bya}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

Figures \ref{fig:expl1} and \ref{fig:expl2} illustrate the difference in the effects of poverty and smoking
on asthma prevalence between genders. The figures highlight the rationale for observing gender 
in studies of the determinants of asthma prevalence -- the adverse
impact of low income appears to be amplified or mitigated by the respondent's gender. Here, the amplification of
poverty effects in female respondents appears to be linear. As Figures \ref{fig:expl3} and \ref{fig:expl4}
demonstrate, however, the interaction of gender with asthma determinants may be more complicated. 
In Figure \ref{fig:expl3},
we observe that female respondents in the *Below high school* education group are more likely to suffer from asthma
compared to female respondents in the *High school* group. Among male respondents, however, the interaction
is reversed. In Figure \ref{fig:expl4}, we can see that female respondents identifying as the *Other* ethnicity
group have the highest asthma prevalence rate compared to other groups -- this observation does not hold for the
male gender.

``` {r echo = FALSE, fig.cap = "Poverty effects by gender\\label{fig:expl1}", fig.pos = "H", fig.align = "center", fig.height = 4}
ggplot(df.p, aes(x = factor(poverty),
                 fill = asthma)) +
    geom_bar(stat = "count",
             position = "fill") +
    facet_wrap(df.p$gender) +
    labs(x = "poverty", y = "proportion")
```

``` {r echo = FALSE, fig.cap = "Smoking effects by gender\\label{fig:expl2}", fig.pos = "H", fig.align = "center", fig.height = 4}
ggplot(df.p, aes(x = factor(smoke),
                 fill = asthma)) +
    geom_bar(stat = "count",
             position = "fill") +
    facet_wrap(df.p$gender) +
    labs(x = "smoking", y = "proportion")
```

``` {r echo = FALSE, fig.cap = "Education effects by gender\\label{fig:expl3}", fig.pos = "H", fig.align = "center", fig.height = 4}
df.p$education <- factor(df.p$education,
                         levels = c("Below high school",
                                    "High school",
                                    "Above high school"))

ggplot(df.p, aes(x = factor(education),
                 fill = asthma)) +
    geom_bar(stat = "count",
             position = "fill") +
    facet_wrap(df.p$gender, nrow = 2) +
    labs(x = "education", y = "proportion") +
    coord_flip()
```

``` {r echo = FALSE, fig.cap = "Ethnicity effects by gender\\label{fig:expl4}", fig.pos = "H", fig.align = "center", fig.height = 4}
ggplot(df.p, aes(x = factor(ethnicity),
                 fill = asthma)) +
    geom_bar(stat = "count",
             position = "fill") +
    facet_wrap(df.p$gender, nrow = 2) +
    labs(x = "ethnicity", y = "proportion") +
    coord_flip()
```

Figure \ref{fig:expl5} shows the distribution of *bmi* in groups by gender and asthma. Although the average
body mass index among male respondents with and without asthma is approximately the same, the distribution of *bmi*
in the asthma group is skewed towards a lower BMI. Female respondent BMI values are more dispersed in general,
with the average BMI in the asthma group being remarkably higher.

``` {r echo = FALSE, fig.cap = "BMI distribution by gender and asthma\\label{fig:expl5}", fig.pos = "H", fig.align = "center"}
ggplot(df.p, aes(x = gender,
                 y = bmi,
                 color = asthma)) +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(),
               size = 1,
               alpha = 0.5)
```

Figure \ref{fig:expl6} reports the distribution of *age* among male and female respondents by *asthma*.
The Figure suggests that women with asthma tend to be younger while men with asthma tend to be older.
Notably, the mean age of the no-asthma group is higher for men. This may be a result of the gender imbalance
introduced in the sample by excluding missing observations.

``` {r echo = FALSE, fig.cap = "BMI distribution by gender and asthma\\label{fig:expl6}", fig.pos = "H", fig.align = "center"}
ggplot(df.p, aes(x = gender,
                 y = age,
                 fill = asthma)) +
    geom_boxplot()
```

Overall, the results of the exploratory analysis support the validity of the primary analysis sample.
The results, further, 
suggest that all socio-demographic variables in the sample affect asthma prevalence, with direction
and magnitude of effects varying in male and female groups.

## Primary Analysis

Table \ref{tbl:prim1} reports the coefficients estimated in the baseline logistic regression model. The baseline
model is fitted on a full sample with no balancing. Year dummies are not included since asthma prevalence does not
appear to be dependent on time based on the results of the exploratory analysis and a model fit with year dummies
included (the raw output and analysis of deviance for two models are reported in 
Appendix \ref{a2}). The AIC of the model in Table \ref{tbl:prim1} is 6655.

*age* is the only variable that is not statistically significant in the baseline model. 
*gender* is statistically significant at the 99% confidence level with the largest associated 
coefficient in terms of magnitude. The coefficient implies that men are 49.52% less likely to have asthma.
Respondents in all ethnicity groups other than *Hispanic* have from 52.77% to 63.86% higher risk of asthma.
Household income below the poverty line increases the risk of asthma by 46.95%. 
The coefficient associated with *Above high school* education is statistically significant at the 95% 
confidence level and suggests that higher education may reduce asthma prevalence by 17.32%. 
*High school* level of education, however, is not significant at the same confidence level.
*smoke* is significant at the 95% confidence level with an associated increase in asthma prevalence of 16.72%.
Finally, a unit increase in BMI, based on the estimate, is associated with a 38.38% increase in the risk of asthma,
significant at the 95% confidence level.

To understand the potential interaction between gender and demographic variables, 
following @greenblattGenderspecificDeterminantsAsthma2017, 
I support the baseline analysis with separate models in *Male* and *Female* groups.
Tables \ref{tbl:prim2} and \ref{tbl:prim3} in Appendix \ref{a1} report the estimated coefficients.
In the *Female* group, the *BMI* differences appear to have a much more substantial contribution to asthma
prevalence (41.61% risk increase versus 31.15% in the *Male* group). 
*education* does not appear to have a significant effect on asthma prevalence among male respondents. 
In the *Female* group, higher education at least at the high school level reduces the risk of asthma by 24.42%,
significant at the 95% confidence level.
The *Male* group, in contrast, has much larger *ethnicity* and *poverty* effects.
The coefficient associated with smoking is not statistically significant in either group.
*age* has a weak (12.77%) effect in the *Male* group and no significant effects in the *Female* group.

``` {r echo = FALSE, message = FALSE, warning = FALSE}
df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

## One-hot encode nominal variables
x <- df %>% select(-c(asthma, year))
x <- dummyVars("~ .", x, fullRank = TRUE)
x <- data.frame(predict(x, newdata = df))

## Recode response as binary
y <- ifelse(df.p$asthma == "Yes", 1, 0)

## Baseline logit
baseline.noy <- glm(y ~ ., cbind(x, y),
                family = "binomial")

baseline.noy %>%
    tidy() %>%
    kable("latex",
          escape = FALSE,
          booktabs = TRUE,
          linesep = "",
          caption="Baseline logistic regression model\\label{tbl:prim1}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

Although the baseline model provides useful inferences, it is not particularly accurate when it comes to predicting
asthma prevalence based on socio-demographic characteristics.
As Figure \ref{fig:prim1} illustrates, the model struggles with identifying respondents who are likely to have
asthma ($AUC = 0.664$). Table \ref{tbl:prim4} reports the confusion matrix with a probability threshold of 0.2.
The true positive rate at this threshold is only 17.84%. At the threshold of 0.5, the model predicts that only
3 of the respondents have asthma.

``` {r echo = FALSE, message = FALSE, warning = FALSE}
df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

## One-hot encode nominal variables
x <- df %>% select(-c(asthma, year))
x <- dummyVars("~ .", x, fullRank = TRUE)
x <- data.frame(predict(x, newdata = df))

## Recode response as binary
y <- ifelse(df.p$asthma == "Yes", 1, 0)

baseline.noy <- glm(y ~ ., cbind(x, y),
                    family = "binomial")

df$prob <- predict(baseline.noy,
                   newdata = cbind(x, y),
                   type = "response")
df$class <- ifelse(df$prob < 0.2, 0, 1)
df$class <- factor(df$class,
                   levels = c(0, 1),
                   labels = c("Never", "Yes"))

confm <- df %>%
    conf_mat(truth = asthma, class)

kable(confm$table,
      "latex",
      escape = FALSE,
      booktabs = TRUE,
      linesep = "",
      caption="Baseline confusion matrix\\label{tbl:prim4}") %>%
    add_header_above(c("", "Truth", "Truth")) %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE, results = FALSE, fig.cap = "Baseline ROC curve\\label{fig:prim1}", fig.pos = "H", fig.align = "center", fig.height = 6, fig.width = 6}
df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

## One-hot encode nominal variables
x <- df %>% select(-c(asthma, year))
x <- dummyVars("~ .", x, fullRank = TRUE)
x <- data.frame(predict(x, newdata = df))

## Recode response as binary
y <- ifelse(df.p$asthma == "Yes", 1, 0)

baseline.noy <- glm(y ~ ., cbind(x, y),
                    family = "binomial")

df$prob <- predict(baseline.noy,
                   newdata = cbind(x, y),
                   type = "response")
df$class <- ifelse(df$prob < 0.2, 0, 1)
df$class <- factor(df$class,
                   levels = c(0, 1),
                   labels = c("Never", "Yes"))

roc.curve(df$asthma, df$prob)
```

While the quality of predictions may be slightly improved by moving the threshold further down, the main problem is,
indeed, the unbalanced response. Before comparing the results of the logistic regression fit
with a random forest fit, as discussed in \ref{methods}, this study explores whether the model accuracy
can be improved by employing the ROSE upsampling [@lunardonROSEPackageBinary2014].
Predictions produced by a ROSE logistic regression model with 10-fold cross-validation and
optimisation by ROC are presented in Table \ref{tbl:prim5}. 
As is apparent from the Table, the upsampled model is aggressively biased towards correctly identifying
respondents with asthma -- at the 0.5 threshold,
the true positive rate in the testing sample is 60%. 
The precision of the model in the testing sample, however, is 16.95% showing that the improvement in accuracy
came at a cost of a high false positive rate. 
The trade-off is visualised in \ref{fig:prim2}. The AUC of the model is 0.63.

``` {r echo = FALSE, message = FALSE, warning = FALSE}
df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

x <- df %>% select(-c(asthma, year))
x <- dummyVars("~ .", x, fullRank = TRUE)
x <- data.frame(predict(x, newdata = df))
y <- df$asthma
df <- cbind(x, y)

inTraining <- createDataPartition(df$y, p = .75, list = FALSE)
training <- df[ inTraining,]
testing  <- df[-inTraining,]

trc <- trainControl(method = "repeatedcv",
                    number = 10,
                    repeats = 10,
                    classProbs = TRUE,
                    summaryFunction = twoClassSummary,
                    search = "random")

tr.bal <- ROSE(y ~., data = training)$data

log.ROSE <- train(y ~ .,
                  data = tr.bal,
                  method = "glm",
                  metric = "ROC",
                  trControl = trc,
                  family = "binomial")

## Scores
tr.bal$prob <- predict(log.ROSE,
                       newdata = tr.bal,
                       type = "prob")
tr.bal$prob <- tr.bal$prob$Yes
tr.bal$class <- predict(log.ROSE,
                        newdata = tr.bal)
tr.bal$data <- "train"

testing$prob <- predict(log.ROSE,
                        newdata = testing,
                        type = "prob")
testing$prob <- testing$prob$Yes
testing$class <- predict(log.ROSE,
                         newdata = testing)
testing$data <- "test"

confm <- testing %>%
    conf_mat(truth = y, class)

kable(confm$table,
      "latex",
      escape = FALSE,
      booktabs = TRUE,
      linesep = "",
      caption="ROSE confusion matrix\\label{tbl:prim5}") %>%
    add_header_above(c("", "Truth", "Truth")) %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE, results = FALSE, fig.cap = "ROSE ROC curve\\label{fig:prim2}", fig.pos = "H", fig.align = "center", fig.height = 6, fig.width = 6}
roc.curve(testing$y, testing$prob)
```

Whether this trade-off is desirable in the context of health care is difficult to determine.
On one hand, identifying as many asthma cases as possible is preferable. On the other hand, incorrect diagnoses
incur social costs. In any case, the ROSE logistic regression fitted in this study is inappropriate for two reasons.
First, the increase in sensitivity is rather small compared to the drastic reduction in specificity.
Second, as the raw output of the ROSE model in Appendix \ref{a2} shows,
higher accuracy is achieved by simply ignoring the effects of predictors and boosting the base probability
of asthma -- the intercept term.

The confusion matrix for a random forest model fitted on the raw sample with 10-fold cross-validation
is presented in Table \ref{tbl:prim6}. 
As we can see, the random forest algorithm, likewise, struggles with the unbalanced response -- all respondents
at the 0.5 threshold are classified as not having asthma. 
Interestingly, as shown in Table \ref{tbl:prim7}, the fitted model considers *age* an extremely important
variable for determining asthma prevalence.

``` {r echo = FALSE, message = FALSE, warning = FALSE}
df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

x <- df %>% select(-c(asthma, year))
x <- dummyVars("~ .", x, fullRank = FALSE)
x <- data.frame(predict(x, newdata = df))
y <- df$asthma
df <- cbind(x, y)

inTraining <- createDataPartition(df$y, p = .75, list = FALSE)
training <- df[ inTraining,]
testing  <- df[-inTraining,]

trc <- trainControl(method = "cv",
                    number = 10,
                    classProbs = TRUE,
                    summaryFunction = twoClassSummary,
                    search = "grid")

tune.grid <- expand.grid(mtry = c(2:10))

rf.fit <- train(y ~ .,
                data = training,
                method = "rf",
                metric = "ROC",
                trControl = trc,
                tuneGrid = tune.grid)

testing$prob <- predict(rf.fit,
                        newdata = testing,
                        type = "prob")
testing$prob <- testing$prob$Yes
testing$class <- predict(rf.fit,
                         newdata = testing)
testing$data <- "test"


confm <- testing %>%
    conf_mat(truth = y, class)

kable(confm$table,
      "latex",
      escape = FALSE,
      booktabs = TRUE,
      linesep = "",
      caption="RF confusion matrix\\label{tbl:prim6}") %>%
    add_header_above(c("", "Truth", "Truth")) %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

The confusion matrix for a random forest model fitted on the ROSE upsampled training sample with 10-fold 
cross-validation is presented in Table \ref{tbl:prim8}. The model has a higher precision compared to the
ROSE-upsampled logistic regression model (24.47%) but does not reach the same true positive rate -- the rate in
the random forest model is only 21.48%. More importantly, however, as demonstrated in Table \ref{tbl:prim9},
the upsampled model seems to correctly identify *gender*, *poverty*, and *ethnicity* as the most important
variables. This validates the baseline logistic regression results and suggests that the model can be further
tuned to improve accuracy.

``` {r echo = FALSE, message = FALSE, warning = FALSE}
imp <- varImp(rf.fit)

kable(imp$importance,
      "latex",
      escape = FALSE,
      booktabs = TRUE,
      linesep = "",
      caption="Variable importance\\label{tbl:prim7}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE}

df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

x <- df %>% select(-c(asthma, year))
x <- dummyVars("~ .", x, fullRank = FALSE)
x <- data.frame(predict(x, newdata = df))
y <- df$asthma
df <- cbind(x, y)

inTraining <- createDataPartition(df$y, p = .75, list = FALSE)
training <- df[ inTraining,]
testing  <- df[-inTraining,]

trc <- trainControl(method = "cv",
                    number = 10,
                    classProbs = TRUE,
                    summaryFunction = twoClassSummary,
                    search = "grid")

tune.grid <- expand.grid(mtry = c(2:10))
training <- ROSE(y ~., data = training)$data

rf.ROSE <- train(y ~ .,
                 data = training,
                 method = "rf",
                 metric = "ROC",
                 trControl = trc,
                 tuneGrid = tune.grid)

testing$prob <- predict(rf.ROSE,
                        newdata = testing,
                        type = "prob")
testing$prob <- testing$prob$Yes
testing$class <- predict(rf.ROSE,
                         newdata = testing)
testing$data <- "test"

confm <- testing %>%
    conf_mat(truth = y, class)

kable(confm$table,
      "latex",
      escape = FALSE,
      booktabs = TRUE,
      linesep = "",
      caption="ROSE RF confusion matrix\\label{tbl:prim8}") %>%
    add_header_above(c("", "Truth", "Truth")) %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE}
imp <- varImp(rf.ROSE)

kable(imp$importance,
      "latex",
      escape = FALSE,
      booktabs = TRUE,
      linesep = "",
      caption="Variable importance (ROSE)\\label{tbl:prim9}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

\newpage

# Conclusion{#conc}

The study finds that *gender*, *poverty*, *ethnicity*, and *bmi* are the most significant socio-demographic variables
contributing to asthma prevalence. 
*poverty* and *ethnicity*, contrary to the findings in @greenblattGenderspecificDeterminantsAsthma2017,
are significant in both gender groups, with a particularly strong impact on male respondents.
*bmi*, conversely, has stronger effects on the risk of asthma among female respondents. *age* and *smoke* have weak
but incosistent effects on asthma prevalence. *Below high school* education appears to negatively affect
asthma prevalence rates in women. 
The effects of all socio-demographic factors vary in gender groups and may be non-linear.

The difficulties with fitting a model with high true positive and low false positive rates
highlight the methodological issues that arise in investigations of asthma determinants. 
As @dharmageEpidemiologyAsthmaChildren2019 emphasise,
the complex nature of asthma and its causes pose a significant challenge for asthma diagnoses and
statistical investigations based on diagnosis data. 
NHANES surveys have been an extremely useful resource for asthma research in the US 
[@greenblattGenderspecificDeterminantsAsthma2017; @hanAdiposityFractionalExhaled2014]; however, the complex design
of the surveys, the lack of laboratory analyses, and the relatively small per-cycle
sample sizes make statistical examinations of the NHANES data a challenging task. This study makes several 
assumptions that are difficult to validate empirically. In particular, it assumes that not only the
effects of potential asthma determinants but also the 
socio-demographic variables that carry these effects are stable across time.
This assumption is rather strong and, indeed, some contradictions between the results
in this paper and the results in @greenblattGenderspecificDeterminantsAsthma2017 can potentially be attributed to
the fact that this study used 5 NHANES survey cycles instead of 3.

Another methodological assumption made in this study is that the unweighted and combined sample of NHANES data
is valid for investigations of asthma. Using this sample without considering survey design essentially makes
the research question of the study
*"Are there gender-specific socio-demographic determinants of asthma in general?"* instead of
*"Are there gender-specific socio-demographic determinants of asthma in the US?"*.
While the differences between primary and survey samples have been explored in the study and 
found to be negligible, it is recommended to not use the results for making population inferences without 
further and more in-depth validation of the assumption. This is, perhaps,
the most severe limitation of the study, especially considering that in-sample bias may make it difficult
to validate the results using another data source. Nonetheless, replicating and validating the results
of this study and @greenblattGenderspecificDeterminantsAsthma2017 is a potential avenue for future research.

A final limitation of this study is associated with the statistical techniques employed. 
The analyses in \ref{analysis}, particularly the exploratory analysis, illustrate vividly that
the interactions between socio-demographic determinants of health outcomes are complex and the approaches
to statistical modelling must be carefully considered. Ultimately, this has led to the decision of observing
the unweighted NHANES sample in this study -- with a view to apply more flexible techniques. Another potential
approach to untangling the asthma determinant interaction complexity is specifying models that explicitly
address the interactions.

\newpage

# References{-}
<div id="refs"></div>

\newpage 

\renewcommand{\thesection}{\Alph{section}}
\setcounter{section}{1}

# Appendix{-}

## Supporting Tables{#a1}

``` {r echo = FALSE}
vars <- data.frame("Variable" = c("gender",
                                  "age",
                                  "ethnicity",
                                  "poverty",
                                  "education",
                                  "bmi",
                                  "smoke",
                                  "year",
                                  "asthma"),
                   "NHANES" = c("RIAGENDR",
                                "RIDAGEYR",
                                "RIDRETH1",
                                "INDFMPIR",
                                "DMDEDUC2",
                                "BMXBMI",
                                "SMQ040",
                                "-",
                                "MCQ035"),
                   "Description" = c("Gender of the sample person",
                                     "Age in years at screening",
                                     "Race/ethnicity",
                                     "Ratio of family income to poverty threshold",
                                     "Education level",
                                     "Body Mass Index (kg/m**2)",
                                     "Current smoking status",
                                     "Survey cycle year",
                                     "Current asthma status"))

kable(vars,
      "latex",
      escape = FALSE,
      booktabs = TRUE,
      linesep = "",
      caption="Primary analysis variables\\label{tbl:vars}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE}
bygv <- CreateTableOne(data = df.v, includeNA = TRUE, 
                       strata = "gender", test = FALSE,
                       var = c("age", "ethnicity", "poverty",
                               "education", "bmi", "smoke",
                               "asthma"))

kableone(bygv,
         format = "latex",
         escape = TRUE,
         booktabs = TRUE,
         longtable = TRUE,
         caption="Variable frequencies by gender (full sample)\\label{tbl:bygv}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE}
df.p$year <- as.factor(df.p$year)
byy <- CreateTableOne(data = df.p, includeNA = TRUE,
                       strata = c("asthma", "gender"), test = FALSE,
                       var = c("year"), addOverall = TRUE)

kableone(byy,
         format = "latex",
         escape = TRUE,
         booktabs = TRUE,
         caption="Asthma prevalence by year and gender (primary sample)\\label{tbl:byy}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE}
df.v$year <- as.factor(df.v$year)
byyv <- CreateTableOne(data = df.v, includeNA = TRUE,
                       strata = c("asthma", "gender"), test = FALSE,
                       var = c("year"), addOverall = TRUE)

kableone(byyv,
         format = "latex",
         escape = TRUE,
         booktabs = TRUE,
         caption="Asthma prevalence by year and gender (full sample)\\label{tbl:byyv}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE}
byav <- CreateTableOne(data = df.v, includeNA = FALSE, 
                       strata = "asthma", test = FALSE,
                       var = c("gender", "age", "ethnicity", "poverty",
                               "education", "bmi", "smoke"))

kableone(byav,
         format = "latex",
         escape = TRUE,
         booktabs = TRUE,
         caption="Demographic variables by asthma (full sample)\\label{tbl:byav}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE}
df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

## Male only
df <- df %>%
    filter(gender == "Male") %>%
    select(-c(year, gender))

## One-hot encode nominal variables
x <- df %>% select(-c(asthma))
x <- dummyVars("~ .", x, fullRank = TRUE)
x <- data.frame(predict(x, newdata = df))

## Recode response as binary
y <- ifelse(df$asthma == "Yes", 1, 0)

baseline.m <- glm(y ~ ., cbind(x, y),
                    family = "binomial")

baseline.m %>%
    tidy() %>%
    kable("latex",
          escape = FALSE,
          booktabs = TRUE,
          linesep = "",
          caption="Baseline logistic regression model (gender = Male)\\label{tbl:prim2}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

``` {r echo = FALSE, message = FALSE, warning = FALSE}
df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

## Female only
df <- df %>%
    filter(gender == "Female") %>%
    select(-c(year, gender))

## One-hot encode nominal variables
x <- df %>% select(-c(asthma))
x <- dummyVars("~ .", x, fullRank = TRUE)
x <- data.frame(predict(x, newdata = df))

## Recode response as binary
y <- ifelse(df$asthma == "Yes", 1, 0)

baseline.f <- glm(y ~ ., cbind(x, y),
                    family = "binomial")

baseline.f %>%
    tidy() %>%
    kable("latex",
          escape = FALSE,
          booktabs = TRUE,
          linesep = "",
          caption="Baseline logistic regression model (gender = Female)\\label{tbl:prim3}") %>%
    kable_styling(latex_options = c("HOLD_position"),
                  font_size = 11)
```

\newpage

## Raw Model Output{#a2}

\footnotesize
``` {r}
## Baseline model with year dummies
df <- df.p %>% mutate_at(c("bmi", "age"),
                         ~(scale(.) %>% as.vector))

## Year as factor
df$year <- factor(df$year)

## One-hot encode nominal variables
x <- df %>% select(-c(asthma))
x <- dummyVars("~ .", x, fullRank = TRUE)
x <- data.frame(predict(x, newdata = df))

## Recode response as binary
y <- ifelse(df.p$asthma == "Yes", 1, 0)

## Baseline logit
baseline <- glm(y ~ ., cbind(x, y),
                family = "binomial")

summary(baseline)

## No year
x <- df %>% select(-c(asthma, year))
x <- dummyVars("~ .", x, fullRank = TRUE)
x <- data.frame(predict(x, newdata = df))

baseline.noy <- glm(y ~ ., cbind(x, y),
                    family = "binomial")

summary(baseline.noy)

## Year-noyear comparison
anova(baseline, baseline.noy)
```
\normalsize

\footnotesize
``` {r}
## ROSE resampling with cross-validation
summary(log.ROSE)
```
\normalsize

